---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2021"
subtitle: "Assignment 3 - Due date 02/15/21"
author: "Stefan Chen"
output: pdf_document
geometry: margin=1.2cm

---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo=FALSE, fig.width=8, fig.height=5, tidy.opts=list(width.cutoff=80), message=FALSE, warning=FALSE, tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A01_Sp21.Rmd"). Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2021 Monthly Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r include=FALSE}
#Load/install required package here
library(tseries)
library(forecast)
library(dplyr)
library(readxl)
library(lubridate)
library(ggplot2)
```

```{r include=FALSE}
#Create data frame with the selected column
setwd("/Users/stefanchen/Documents/Duke/Classes/Spring 2021/ENV 790/GitHub/ENV790_30_TSA_S2021/Data")

#Read data
re_select0<-read_xlsx("Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", skip=12, col_names=FALSE)

#Select columns
re_select<-as.data.frame(re_select0[,c(1,4,5,6)]) #Transform to dataframe
colnames(re_select)=c("Date",
                      "Biomass Energy Production",
                      "Renewable Energy Production",
                      "Hydroelectric Consumption") 

#Checking data
head(re_select)
str(re_select)

#Transform dataframe to time series
ts_select<-ts(re_select[,2:4], start=c(1973,1), frequency=12)

my_date<-re_select$Date  #Establish my_date
nobs<-nrow(re_select) #Numbers of rows in ts_select
t<-c(1:nobs) #Set a vector for number of rows in ts_select
```

\pagebreak

## Trend Component

### Q1
**Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: watch videos for M4)**

<br />

```{r fig.width=12, fig.height=6}
#Acf, Pacf, and time series
for(i in 1:3){
  par(mfrow=c(1,3))  #Place plot side by side
  acf(ts_select[,i], lag.max = 40, main=colnames(ts_select)[i])
  pacf(ts_select[,i],lag.max=40, main=colnames(ts_select)[i])
  plot(ts_select[,i], 
     main=colnames(ts_select)[i], 
     ylab="Trillion Btu",
     type="l")
}
```

<br />

### Q2
**From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?**

<br />

Total biomass energy production and total renewable energy production appear to show a growing trend between 1973 and 2020. However, both energy production appear to be stagnant between 1985 and 2000. On the other hand, hydroelectric power consumption shows a slight decreasing trend. 

<br />

### Q3
**Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.**
 
<br />

Summary of linear model: biomass energy production 

```{r}
#Biomass linear model
lmodel_bio<-lm(ts_select[,1]~t)
summary(lmodel_bio)
beta0_bio=as.numeric(lmodel_bio$coefficients[1])  #First coefficient is the intercept term or beta0
beta1_bio=as.numeric(lmodel_bio$coefficients[2])  #Second coefficient is the slope or beta1
```

Summary of linear model: renewable energy production 

```{r}
#Renewable energy linear model
lmodel_re<-lm(ts_select[,2]~t)
summary(lmodel_re)
beta0_re=as.numeric(lmodel_re$coefficients[1]) 
beta1_re=as.numeric(lmodel_re$coefficients[2]) 
```

Summary of linear model: renewable energy production 

```{r}
#Hydropower linear model
lmodel_hydro<-lm(ts_select[,3]~t)
summary(lmodel_hydro)
beta0_hydro=as.numeric(lmodel_hydro$coefficients[1])  
beta1_hydro=as.numeric(lmodel_hydro$coefficients[2])
```

<br />

### Q4
**Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?**

<br />

New plots eliminated the upward trends in biomass and renewable energy production, and the downward trend in hydroelectricity consumption. 

```{r}
#Remove the trend from series
detrend_bio<-re_select[,2]-(beta0_bio+beta1_bio*t)

#Comparison plot for biomass
ggplot(re_select, aes(x=my_date, y=re_select[,2])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title="Biomass Energy Production") + 
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_bio), col="green")+
            geom_smooth(aes(y=detrend_bio),color="orange",method="lm") 
```

```{r}
#Remove the trend from series
detrend_re<-re_select[,3]-(beta0_re+beta1_re*t)

#Comparison plot for biomass
ggplot(re_select, aes(x=my_date, y=re_select[,3])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title="Renewable Energy Production") + 
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_re), col="green")+
            geom_smooth(aes(y=detrend_re),color="orange",method="lm") 
```

```{r}
#Remove the trend from series
detrend_hydro<-re_select[,4]-(beta0_hydro+beta1_hydro*t)

#Comparison plot for biomass
ggplot(re_select, aes(x=my_date, y=re_select[,4])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title="Hydroelectricity Consumption") + 
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_hydro), col="green")+
            geom_smooth(aes(y=detrend_hydro),color="orange",method="lm") 
```

<br />

### Q5
**Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?**

<br />

The autoregression of biomass and renewable energy production appears to fluctuate slightly more than the original autoregression plot although still not passing autoregression of zero.

```{r}
detrend<-data.frame("Biomass Energy Production"=detrend_bio, 
                    "Renewable Energy Production"=detrend_re, 
                    "Hydroelectricity Consumption"=detrend_hydro) #Combine all the detrend model

ts_detrend<-ts(detrend, start=c(1973,1), frequency=12) #time series of the detrend models

#Acf, Pacf, and time series
for(i in 1:3){
  par(mfrow=c(1,2))  #Place plot side by side
  acf(ts_detrend[,i], lag.max=40, main=colnames(ts_detrend)[i])
  pacf(ts_detrend[,i], lag.max=40, main=colnames(ts_detrend)[i])
}
```

<br />

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
**Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.**

<br />

**Biomass Energy Production**

Based on the regression summary of biomass energy production, only coefficient for December appears to show statistically significant effect on the model. Therefore, it can be concluded that there's no significant seasonal trend.

```{r}
#Use seasonal means model
dummies<-seasonaldummy(ts_select[,1])  #Create the seasonal dummies

#Fit a linear model to the seasonal dummies - bio
sm_model_bio<-lm(re_select[,2]~dummies)
summary(sm_model_bio)
```

**Renewable Energy Production**

Based on the regression summary of renewable energy production, only coefficient for December appears to show statistically significant effect on the model. Therefore, it can be concluded that there's no significant seasonal trend.

```{r}
#Fit a linear model to the seasonal dummies - re
sm_model_re<-lm(re_select[,3]~dummies)
summary(sm_model_re)
```

**Hydroelectricity Consumption**

Based on the regression summary of hdroelectricity consumption, nine out of twelve months appears to show statistically significant effect on the model. Therefore, it can be concluded that there is a significant seasonal trend.

```{r}
#Fit a linear model to the seasonal dummies - hydro
sm_model_hydro<-lm(re_select[,4]~dummies)
summary(sm_model_hydro)
```

```{r}
#Store regression coefficients
sm_int_bio<-sm_model_bio$coefficients[1]
sm_int_re<-sm_model_re$coefficients[1]
sm_int_hydro<-sm_model_hydro$coefficients[1]
sm_coeff_bio<-sm_model_bio$coefficients[2:12]
sm_coeff_re<-sm_model_re$coefficients[2:12]
sm_coeff_hydro<-sm_model_hydro$coefficients[2:12]

#Create regression coefficient dataframes
sm_int<-data.frame("bio"=sm_int_bio, "re"=sm_int_re, "hydro"=sm_int_hydro)
sm_coeff<-data.frame("bio"=sm_int_bio, "re"=sm_int_re, "hydro"=sm_int_hydro)
```

<br />

### Q7
**Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?**

<br />

Overall flunctuation appears to be smoothened for all plots. However, the general upward and downward trend observed from the original plots remain observale.

```{r}
#Compute seasonal component
seas_comp_bio<-array(0,nobs) #bio
for(i in 1:nobs){
  seas_comp_bio[i]=(sm_int_bio+sm_coeff_bio%*%dummies[i,])}

seas_comp_re<-array(0,nobs) #re
for(i in 1:nobs){
  seas_comp_re[i]=(sm_int_re+sm_coeff_re%*%dummies[i,])}

seas_comp_hydro<-array(0,nobs) #hydro
for(i in 1:nobs){
  seas_comp_hydro[i]=(sm_int_hydro+sm_coeff_hydro%*%dummies[i,])}

#Dataframe of seas_comp
seas_comp<-data.frame("bio"=seas_comp_bio, "re"=seas_comp_re, "hydro"=seas_comp_hydro)

#Removing seasonal component
deseason_bio<-re_select[,2]-seas_comp[1]
deseason_re<-re_select[,3]-seas_comp[2]
deseason_hydro<-re_select[,4]-seas_comp[3]

#Dataframe of deseasoned
deseason<-data.frame("bio"=deseason_bio, "re"=deseason_re, "hydro"=deseason_hydro)

#Plotting comparison graph
 for(n in 1:3){
   print(
     ggplot(re_select, aes(x=my_date, y=re_select[,n+1])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title=colnames(re_select)[n+1]) +
            geom_line(aes(y=deseason[,n]), col="green")
   )
 }
```

<br />

### Q8
**Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?**

<br />

The autoregression of biomass and renewable energy production appears to fluctuate slightly more than the original autoregression plot although still not passing autoregression of zero.

```{r}
#Create time series of deseasoned model
ts_deseason<-ts(deseason, start=c(1973,1), frequency=12)

#Acf and Pacf
for(i in 1:3){
  par(mfrow=c(1,2))  #Place plot side by side
  acf(ts_detrend[,i], lag.max=40, main=colnames(ts_detrend)[i])
  pacf(ts_detrend[,i],lag.max=40, main=colnames(ts_detrend)[i])
}
```

\pagebreak

## Appendix

```{r echo=TRUE, results='hide', warning=FALSE, message=FALSE, error=FALSE, fig.show='hide'}
knitr::opts_chunk$set(echo=FALSE, fig.width=8, fig.height=6, tidy.opts=list(width.cutoff=80), message=FALSE, warning=FALSE, tidy=FALSE) 

#Load/install required package here
library(tseries)
library(forecast)
library(dplyr)
library(readxl)
library(lubridate)
library(ggplot2)

#Create data frame with the selected column
setwd("/Users/stefanchen/Documents/Duke/Classes/Spring 2021/ENV 790/GitHub/ENV790_30_TSA_S2021/Data")

#Read data
re_select0<-read_xlsx("Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", skip=12, col_names=FALSE)

#Select columns
re_select<-as.data.frame(re_select0[,c(1,4,5,6)]) #Transform to dataframe
colnames(re_select)=c("Date",
                      "Biomass Energy Production",
                      "Renewable Energy Production",
                      "Hydroelectric Consumption") 

#Checking data
head(re_select)
str(re_select)

#Transform dataframe to time series
ts_select<-ts(re_select[,2:4], start=c(1973,1), frequency=12)
ts_select

my_date<-re_select$Date  #Establish my_date
nobs<-nrow(re_select) #Numbers of rows in ts_select
t<-c(1:nobs) #Set a vector for number of rows in ts_select

#Acf, Pacf, and time series
for(i in 1:3){
  par(mfrow=c(1,3))  #Place plot side by side
  acf(ts_select[,i], lag.max = 40, main=colnames(ts_select)[i])
  pacf(ts_select[,i],lag.max=40, main=colnames(ts_select)[i])
  plot(ts_select[,i], 
     main=colnames(ts_select)[i], 
     ylab="Trillion Btu",
     type="l")
}

#Biomass linear model
lmodel_bio<-lm(ts_select[,1]~t)
summary(lmodel_bio)
beta0_bio=as.numeric(lmodel_bio$coefficients[1])  #First coefficient is the intercept term or beta0
beta1_bio=as.numeric(lmodel_bio$coefficients[2])  #Second coefficient is the slope or beta1

#Renewable energy linear model
lmodel_re<-lm(ts_select[,2]~t)
summary(lmodel_re)
beta0_re=as.numeric(lmodel_re$coefficients[1]) 
beta1_re=as.numeric(lmodel_re$coefficients[2]) 

#Hydropower linear model
lmodel_hydro<-lm(ts_select[,3]~t)
summary(lmodel_hydro)
beta0_hydro=as.numeric(lmodel_hydro$coefficients[1])  
beta1_hydro=as.numeric(lmodel_hydro$coefficients[2])

#Remove the trend from series
detrend_bio<-re_select[,2]-(beta0_bio+beta1_bio*t)

#Comparison plot for biomass
ggplot(re_select, aes(x=my_date, y=re_select[,2])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title="Biomass Energy Production") + 
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_bio), col="green")+
            geom_smooth(aes(y=detrend_bio),color="orange",method="lm") 

#Remove the trend from series
detrend_re<-re_select[,3]-(beta0_re+beta1_re*t)

#Comparison plot for biomass
ggplot(re_select, aes(x=my_date, y=re_select[,3])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title="Renewable Energy Production") + 
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_re), col="green")+
            geom_smooth(aes(y=detrend_re),color="orange",method="lm") 

#Remove the trend from series
detrend_hydro<-re_select[,4]-(beta0_hydro+beta1_hydro*t)

#Comparison plot for biomass
ggplot(re_select, aes(x=my_date, y=re_select[,4])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title="Hydroelectricity Consumption") + 
            geom_smooth(color="red",method="lm") +
            geom_line(aes(y=detrend_hydro), col="green")+
            geom_smooth(aes(y=detrend_hydro),color="orange",method="lm") 

detrend<-data.frame("Biomass Energy Production"=detrend_bio, 
                    "Renewable Energy Production"=detrend_re, 
                    "Hydroelectricity Consumption"=detrend_hydro) #Combine all the detrend model

ts_detrend<-ts(detrend, start=c(1973,1), frequency=12) #time series of the detrend models

#Acf, Pacf, and time series
for(i in 1:3){
  par(mfrow=c(1,2))  #Place plot side by side
  acf(ts_detrend[,i], lag.max=40, main=colnames(ts_detrend)[i])
  pacf(ts_detrend[,i], lag.max=40, main=colnames(ts_detrend)[i])
}

#Use seasonal means model
dummies<-seasonaldummy(ts_select[,1])  #Create the seasonal dummies

#Fit a linear model to the seasonal dummies - bio
sm_model_bio<-lm(re_select[,2]~dummies)
summary(sm_model_bio)

#Fit a linear model to the seasonal dummies - re
sm_model_re<-lm(re_select[,3]~dummies)
summary(sm_model_re)

#Fit a linear model to the seasonal dummies - hydro
sm_model_hydro<-lm(re_select[,4]~dummies)
summary(sm_model_hydro)

#Store regression coefficients
sm_int_bio<-sm_model_bio$coefficients[1]
sm_int_re<-sm_model_re$coefficients[1]
sm_int_hydro<-sm_model_hydro$coefficients[1]
sm_coeff_bio<-sm_model_bio$coefficients[2:12]
sm_coeff_re<-sm_model_re$coefficients[2:12]
sm_coeff_hydro<-sm_model_hydro$coefficients[2:12]

#Create regression coefficient dataframes
sm_int<-data.frame("bio"=sm_int_bio, "re"=sm_int_re, "hydro"=sm_int_hydro)
sm_coeff<-data.frame("bio"=sm_int_bio, "re"=sm_int_re, "hydro"=sm_int_hydro)

#Compute seasonal component
seas_comp_bio<-array(0,nobs) #bio
for(i in 1:nobs){
  seas_comp_bio[i]=(sm_int_bio+sm_coeff_bio%*%dummies[i,])}

seas_comp_re<-array(0,nobs) #re
for(i in 1:nobs){
  seas_comp_re[i]=(sm_int_re+sm_coeff_re%*%dummies[i,])}

seas_comp_hydro<-array(0,nobs) #hydro
for(i in 1:nobs){
  seas_comp_hydro[i]=(sm_int_hydro+sm_coeff_hydro%*%dummies[i,])}

#Dataframe of seas_comp
seas_comp<-data.frame("bio"=seas_comp_bio, "re"=seas_comp_re, "hydro"=seas_comp_hydro)

#Removing seasonal component
deseason_bio<-re_select[,2]-seas_comp[1]
deseason_re<-re_select[,3]-seas_comp[2]
deseason_hydro<-re_select[,4]-seas_comp[3]

#Dataframe of deseasoned
deseason<-data.frame("bio"=deseason_bio, "re"=deseason_re, "hydro"=deseason_hydro)

#Plotting comparison graph
 for(n in 1:3){
   print(
     ggplot(re_select, aes(x=my_date, y=re_select[,n+1])) +
            geom_line(color="blue") +
            labs(x="Time", y="Energy (Trillion Btu)", title=colnames(re_select)[n+1]) +
            geom_line(aes(y=deseason[,n]), col="green")
   )
 }

#Create time series of deseasoned model
ts_deseason<-ts(deseason, start=c(1973,1), frequency=12)

#Acf and Pacf
for(i in 1:3){
  par(mfrow=c(1,2))  #Place plot side by side
  acf(ts_detrend[,i], lag.max=40, main=colnames(ts_detrend)[i])
  pacf(ts_detrend[,i],lag.max=40, main=colnames(ts_detrend)[i])
}

```

